Visualizing Independence Using Extended Association Plots

### 2005-08-02 trunk/vcd/paper/JCGS.Rnw
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/vcd@423 edb9625f-4e0d-4859-8d74-9fd3b1da38cb

\begin{center}
\textbf{\LARGE Residual-based Shadings for Visualizing (Conditional) Independence}

@@

We consider a 2-way contingency table with cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively. For
convenience the number of observations is denoted $n = n_{++}$.

Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
measure of discrepancy
between observed and expected values are the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables.
%% reference needed

@@

>>>>

\begin{center}
\textbf{\LARGE Residual-based Shadings for Visualizing\\ (Conditional) Independence}

@@

We consider a 2-way contingency table with cell frequencies $\{n_{ij}\}$ for
$i = 1, \dots, I$ and $j = 1, \dots, J$ and row and column sums
$n_{i+} = \sum_i n_{ij}$ and $n_{+j} = \sum_j n_{ij}$ respectively.
%% For convenience the number of observations is denoted $n = n_{++}$.
Given an underlying distribution with theoretical cell probabilities
$\pi_{ij}$, the null hypothesis of independence of the two categorical
variables can be formulated as
\begin{equation} \label{eq:H0}
H_0: \; \pi_{ij} \quad = \quad \pi_{i+} \pi_{+j}.
\end{equation}

The expected cell frequencies in this model are
$\hat n_{ij} = n_{i+}n_{+j}/n$. The best known and most used
measure of discrepancy
between observed and expected values are probably the Pearson residuals
\begin{equation} \label{eq:Pearson}
r_{ij} \quad = \quad \frac{n_{ij} - \hat n_{ij}}{\sqrt{\hat n_{ij}}}.
\end{equation}
Therefore, a rather intuitive idea is to reject the null hypothesis when
there are residuals which are too extreme, i.e., not close enough to
zero. The most convenient way to aggregate the $I \times J$ residuals
to one test statistic is their squared sum
\begin{equation} \label{eq:ChiSquare}
X^2 \quad = \quad \sum_{i, j}r_{ij}^2,
\end{equation}
because this is known to have a limiting $\chi^2$ distribution with $(I-1)(J-1)$
degrees of freedom under the null hypothesis.
This is the well-known $\chi^2$ test for independence in 2-way tables.
%% reference needed

@@

\begin{table}[htb]
\begin{center}
\begin{tabular}{rr|rrr}
                   &         & \multicolumn{3}{|l}{\textbf{Improved}} \\
                   &         & None & Some & Marked \\ \hline
\textbf{Treatment} & Placebo &   19 &    7 &      6 \\
                   & Treated &    6 &    5 &     16
\end{tabular}
\caption{\label{tab:art} Treatment and improvement among 59 female
patients with rheumatoid arthritis.}
\end{center}
\end{table}
